Static contract checking for Haskell

### 2011-12-20 paper/paper.tex
Paper: cleaned up / removed many things, started removing the 'a' let-binding.

@@

\note[nc]{
Not true that we only consider full application of functions: the definition of satisfaction for arrow contracts
$x:c_1 \to c_2$ already includes partial applications.  We could add a
special cases for each function arity, e.g.
\begin{eqnarray*}
&&\ctrans{f \in x_1:c_1 \to \cdots \to x_n:c_n \to c}^v \\
&&      :=\forall x_1, \ldots, x_n. \min{\full{f}(x_1, \ldots, x_n)} \\
&&\quad   \to \left( \bigwedge_{i=1}^n \ctrans{x_i \in c_i}^{\bar{v}} \right)
              \to \ctrans{\full{f}(x_1, \ldots, x_n) \in c}^v
\end{eqnarray*}
for functions of arity $n$, to make the claim true, but that would add
more clutter.
}, in order to remove clutter. Dealing with
  partial application is not hard but a bit cumbersome.
\end{itemize}

@@

\subsection{Contracts}
Contract syntax is described in figure \ref{cont-stx}. The predicates
$p$ in contracts range over boolean $\hprime$ expressions. We
only consider pairs of contract for simplicity, although there is no
issue with generalisation to arbitrary tuples.\note[nc]{
%
  XXX: Tuple contracts are not actually supported in the
  implementation. Should they be discussed here?
%
So, ``tuples'' refers to the the term constructor, not the type
constructor, and the generalization is e.g.
\[
e \in Cons~c_1~c_2 \iff e \text{ diverges or } 
      (e \to^\star Cons~e_1~e_2 \mbox{ and } e_1 \in c_1, e_2 \in c_2)
\]
?
}

@@

\subsection{BAD and UNR}
Consider the following piece of code:
\begin{verbatim}
a = 0 + True

b ::: CF
b = undefined

c = error "foo"
\end{verbatim}

\begin{itemize}
\item $a$ is ill-typed
\item $b$'s implementation is not correct wrt its contract
\item $c$ goes through the whole toolchain (compiler, typechecker, contractchecker)
\end{itemize}

One thing to notice is that $a$ and $b$ are things that ``should not
happen'' but are caught statically whereas $c$ should not happen but
can only be dealt with dynamically.\note[nc]{
I find this example confusing: \texttt{error} calls should be just as
easy to find as \texttt{undefined} calls statically.}

We can now define two types of problematic expressions: those that
cannot happen during a run of the program and those that can.
Expressions of the first type are called unreachable (and equated to
the special value $\unr$ in our first-order theory)\note[nc]{
%
In particular, ill-typed expressions are $\unr$, not $\bad$
%
}, whilst
expressions of the latter type are called bad (and equated to the
special value $\bad$).

We said earlier that we only considered syntactically correct and
well-typed programs as input. That implies that the ``$a$'' case
cannot happen. But given that our first-order logic is not typed, the
theorem prover may decide to instanciate a variable with an ill-typed
value! \change[nc]{In order to prevent this, we will need to encode some basic
type-checking mecanism directly in our first-order theory.}{This is OK, because the axioms are designed so that scrutinizing an ill-typed constructor-head results in $\unr$.  On the other hand, some ill-typed expressions are not $\unr$, e.g. we can prove that $tail~(Cons~Nil~Z) = Z$.}

\subsection{First-order logic with equality}
We use first-order logic with equality, defined in figure \ref{fol-stx}.\note[nc]{I guess ``with equality'' means that the proof theory has an axiom schema for substition, i.e. forall $\phi$, $\forall x,y.x=y \to (\phi \tot \phi[y/x])$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.}
\note[nc]{See Section~\ref{sec:contracts} for a discussion of the $\min{}$ predicate.} 
The expression language in our logic is the expression language of 
\hprime (Figure~\ref{fig:hprime-syntax}).  Note that curried application in \hprime
corresponds to a binary function symbol in FOL, which we denote by juxtaposition.

@@

\subsection{$\dtrans{}$ -- Definitions}
We give in figure~\ref{dtrans} the two translations of function
definitions.

figure~\ref{dtrans} gives the translation of function not defined by
pattern matching, which is really easy: we just have to state the
equality between le left-hand side and the right-hand side.

Translating definitions that use pattern-matching is more challenging
and is described in figure~\ref{dtrans}.  The first line says that
when applied to an argument that matches a pattern of the case
expression, we should equate the function call to the corresponding
expression. The second line states that if the pattern-matching failed
or if we pattern-matched on $\bad$ then the result should be
$\unr$.\note[nc]{This text does not agree with the figure.  In
  particular, the axiom says nothing about the result of scrutinizing
  $\bad$.  I think the correct behavior is that scrutinizing $\bad$
  results in $\bad$, \emph{not} $\unr$.}

\begin{figure*}
\[
\begin{array}{rcl}
  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \phi}} \\[2.5mm]
  \dtrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} &=& \ttrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} \\
  \dtrans{f \in c} &=& \ctrans{f \in c}\\
  \dtrans{f~\vec{x} = e} &=& \forall \vec{x},a.~ (\min{a} \land a =
  \full{f}(\vec{x})) \to a = e \\
   \raisebox{ 0.5 cm }{$\dtrans{f~\vec{x} = \mbox{case } e \mbox{ of } [K_i~\vec{x_i} = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
     \forall \vec{x},a.~(\mbox{min}(a) \land a = \full{f}(\vec{x})) &\to&
     \multicolumn{3}{c}{  \texttt{min}(e) \land (e = \texttt{BAD} \lor
       \left(\bigvee_i \exists \vec{x_i}.~e = \full{K}_i(\vec{x_i})\right) \lor a = \texttt{UNR})} \\
     &\land& \forall \vec{x_1}.~e = \full{K}_1(\vec{x_1}) &\to& a = e_1 \\
     &\land& & \cdots & \\
     &\land& \forall \vec{x_n}.~e = \full{K}_n(\vec{x_n}) &\to& a = e_n \\
   \end{array} 
\end{array}
\]
\caption{$\dtrans{}$ -- Defintions translation}
 \label{dtrans}
 \note[nc]{??? Why do we introduce the extra quantified variable $a$,
   representing the result (``answer''?), instead of just using the result directly
   (which is logically equivalent)? Same question for the next
   Figure.  Does this make a difference to Equinox?}
\end{figure*}

@@

\begin{itemize}
\item $(\phi_\text{injective})$ Constructors are injective.  That is, for
  each term constructor $K$,
  if $K e_1 \cdots e_n = K e'_1 \cdots e'_n$, then $e_i = e'_i$ for
  all $i$.  We axiomitize this via projections $\proj{i}{K}$: for each $i$,
  $\proj{i}{K}(K e_1 \cdots e_n) = e_i$.
\item $(\phi_\text{disjoint})$ Distinct constructors have disjoint ranges.
\item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\gets$: if we pack crash-free values in a data constructor,
    the resulting value is crash-free.  For $K$ has arity $n$
    we assert $\ctrans{K \in \cfc^n \to \cfc}^-$.

  \item $\to$: a value $t$ of type $T$\note[nc]{
    ``type'' is used loosely here: it means the constructor head is of type $T$.
    } is crash-free implies
    that every value packed in it is crash-free, since we can use
    a case expression to extract any packed value.
    \note[nc]{XXX: This translation is ad-hoc.}

  Note that this is not true for functions: a function is not required
  to use all of its arguments. \texttt{fst} is crash-free if and only
  if the first argument of the pair is crash-free. The second argument
  is irrelevant.

@@

\item $(\phi_\text{lazy})$ No full application of a constructor $K_i$ is unreachable or crashing.
\item One may want to also state that if for all $j$, $x_j \neq \bad$ then
  $\full{K}_i(\vec{x}) \neq \bad$. It is already implied by the fact that
  $\left(\forall j. \cf{x_j}\right) \to \cf{\full{K}_i(\vec{x})}$.
\end{itemize}

>>>>

  The following argument has major problems:\begin{enumerate}
  \item it's not clear what ``terminating terms'' means, since our
    language is lazy.
  \item consistency of $\trans{}^+$ (checking) does not imply
    consistency of the $\trans{}^-$ (assuming), e.g. because the function
    definitions change (we only use anonymous recursive functions in
    $\trans{}^+$).  Do we need to argue that our induction is sound? We
    can't simply define assumed ($\trans{}^-$) functionss as $\unr$,
    since e.g. $\unr \neq S ~x$ for all $x$.
  \item the presentation of $\cfc$ is wrong: asking if something is
    equal to $\bad$ only works in a strict language.
  \end{enumerate}

@@

  \note[nc]{ UPDATE: actually, this is true now, but we still want to
    talk about collapsing arrows, because the implementation does
    that?

Not true that we only consider full application of functions: 
the definition of satisfaction for arrow contracts
$x:c_1 \to c_2$ already includes partial applications.  We could add a
special cases for each function arity, e.g.
\begin{eqnarray*}
&&\ctrans{f \in x_1:c_1 \to \cdots \to x_n:c_n \to c}^v \\
&&      :=\forall x_1, \ldots, x_n. \min{\full{f}(x_1, \ldots, x_n)} \\
&&\quad   \to \left( \bigwedge_{i=1}^n \ctrans{x_i \in c_i}^{\bar{v}} \right)
              \to \ctrans{\full{f}(x_1, \ldots, x_n) \in c}^v
\end{eqnarray*}
for functions of arity $n$, to make the claim true, but that would add
more clutter.
}, in order to remove clutter. Dealing with
  partial application is not hard but a bit cumbersome.
\end{itemize}

@@

\subsection{BAD and UNR}
There are two types of problematic expressions: those that cannot
happen during a run of the program and those that can.  We call
infinite loops, and expressions of the first type, such as type
errors, \emph{unreachable} and equate them to the special value $\unr$
in our FOL theory.\footnote{
%
In particular, ill-typed expressions are $\unr$, not $\bad$
%
} We call expressions of the second type \emph{bad} and equate them to
the special value $\bad$ in our FOL theory.

Although we only consider well-typed programs, 
our FOL theory is not typed,\footnote{
%
  We could encode types with predicates.  Will Sonnex (Zeno) did this 
  when encoding examples in ACL2s, in order to improve performance.  E.g.
  \[
  \forall x,y. isNat(x) \land isNat(y) \to isNat(add(x,y)) \land \text{ standard stuff}.
  \]
  Things might become more complicated for higher order functions, e.g.
  \begin{align*}
  \forall f,xs,a,b.& isType(a) \land isType(b) \\
                   &\land hasType(f,arrowType(a,b)) \\
                   &\land hasType(xs,listType(a))\\
                   &\to hasType(map(f,xs),listType(b)) \land \text{ standard stuff}.
  \end{align*}
%
}
and so quantifiers in FOL range over \emph{all}\footnote{
%
  Well, technically, in \hprime we have both $f$ and $\full{f}$ for
  each declared function $f$, and the quantifier only includes $f$.
%
} expressions in \hprime.
This is OK, because the axioms are designed so that scrutinizing an
ill-typed constructor-head results in $\unr$.  On the other hand,
some ill-typed expressions are not $\unr$, e.g. we can prove
that $tail~(Cons~Nil~Z) = Z$.\footnote{
%
  Basically, if we discover that an expression is ill-typed, then we return $\unr$, but
  we do this lazily.
%
}

\subsection{Contracts}
Contract syntax is described in figure \ref{cont-stx}. The predicates
$p$ in predicate contracts range over boolean $\hprime$ expressions. We
only consider pairs of contract for simplicity, although there is no
issue with generalisation to arbitrary tuples.\note[nc]{
%
  XXX: Tuple contracts are not actually supported in the
  implementation. Should they be discussed here?
%
So, ``tuples'' refers to the the term constructor, not the type
constructor, and the generalization is e.g.
\[
e \in Cons~c_1~c_2 \iff e \text{ diverges or } 
      (e \to^\star Cons~e_1~e_2 \mbox{ and } e_1 \in c_1, e_2 \in c_2)
\]
?
}

@@

\subsection{First-order logic with equality}
We use first-order logic with equality, defined in figure \ref{fol-stx}.\footnote{
%
I guess ``with equality'' means that the proof theory has an axiom schema for substition, i.e. forall $\phi$, $\forall x,y.x=y \to (\phi \tot \phi[y/x])$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.
%
}
The expression language in our logic is the expression language of 
\hprime (Figure~\ref{fig:hprime-syntax}).  Note that curried application in \hprime
corresponds to a binary function symbol in FOL, which we denote by juxtaposition.

@@

\subsection{$\dtrans{}$ -- Definitions}
We give in figure~\ref{dtrans} the translation of function
definitions. The first line says that when applied to an argument that
matches a pattern of the case expression, we should equate the
function call to the corresponding expression. The second line states
that if the pattern-matching failed or if we pattern-matched on $\bad$
then the result should be $\unr$.\note[nc]{This text does not agree
  with the figure.  In particular, the axiom says nothing about the
  result of scrutinizing $\bad$.  I think the correct behavior is that
  scrutinizing $\bad$ results in $\bad$, \emph{not} $\unr$.}

\newcommand{\fullfxs}{\full{f}(\vec{x})}
\begin{figure*}
\[
\begin{array}{rcl}
  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \phi}} \\[2.5mm]
  \dtrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} &=& \ttrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} \\
  \dtrans{f \in c} &=& \ctrans{f \in c}\\
  \dtrans{f~\vec{x} = e} &=& \forall \vec{x}.\min{\fullfxs} 
                                              \to \fullfxs = e \\
   \raisebox{ 0.5 cm }{$\dtrans{f~\vec{x} = \mbox{case } e \mbox{ of } [K_i~\vec{x_i} = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
     \forall \vec{x}. \min{\fullfxs} &\to&
     \multicolumn{3}{c}{  \min{e} \land (e = \texttt{BAD} \lor
       \left(\bigvee_i \exists \vec{x_i}.~e = \full{K}_i(\vec{x_i})\right) \lor \fullfxs = \texttt{UNR})} \\
     &\land& \forall \vec{x_1}.~e = \full{K}_1(\vec{x_1}) &\to& \fullfxs = e_1 \\
     &\land& & \cdots & \\
     &\land& \forall \vec{x_n}.~e = \full{K}_n(\vec{x_n}) &\to& \fullfxs = e_n \\
   \end{array} 
\end{array}
\]
\caption{$\dtrans{}$ -- Defintions translation}
 \label{dtrans}
 \note[nc]{??? Why do we introduce the extra quantified variable $a$,
   representing the result (``answer''?), instead of just using the result directly
   (which is logically equivalent)? Same question for the next
   Figure.  Does this make a difference to Equinox?}
\end{figure*}

@@

\begin{itemize}
\item $(\phi_\text{injective})$ Constructors are injective.  That is, for
  each term constructor $K$,
  if $K e_1 \cdots e_n = K e'_1 \cdots e'_n$, then $e_i = e'_i$ for
  all $i$.  We axiomitize this via projections $\proj{i}{K}$: for each $i$,
  $\proj{i}{K}(K e_1 \cdots e_n) = e_i$.
\item $(\phi_\text{disjoint})$ Distinct constructors have disjoint ranges.
\item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\gets$: if we pack crash-free values in a data constructor,
    the resulting value is crash-free.  If $K$ has arity $n$
    then we assert $\ctrans{K \in \cfc^n \to \cfc}^-$.

  \item $\to$: a value $t$ of type $T$\footnote{
    ``type'' is used loosely here: it means the constructor 
      head is from the type $T$.
    } is crash-free implies
    that every value packed in it is crash-free, since we can use
    a case expression to extract any packed value.
    \note[nc]{XXX: This translation is ad-hoc.}

  Note that this is not true for functions: a function is not required
  to use all of its arguments. For example, \texttt{fst} is crash-free if and only
  if the first argument of the pair is crash-free. The second argument
  is irrelevant.

@@

\item $(\phi_\text{lazy})$ No full application of a constructor
  $K_i$ is unreachable or crashing.
\end{itemize}
